---
title: "Your first linear regression"
author: "Bioinformatics Learning Workspace"
date: "1/15/2022"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
---

\newpage

# Calculating linear regression (so easy you can do it by hand)

## Formula for calculating the best slope for a set of points.

$$m=\frac{\overline{xy}-(\overline{x})(\overline{y})}{\overline{x^{2}}-(\overline{x})^{2}}$$

## Formula for calculating the y-intercept for a set of points

$$b=\overline{y} - m\overline{x}$$

## Example 

```{r regression_data, echo=FALSE}

x <- c(1,2,3,5)
y <- c(2,3,4,5)

my_df <- data.frame(x=x,y=y) 
knitr::kable(my_df)


plot(x,y)
```

So, for 
$$m=\frac{\overline{xy}-(\overline{x})(\overline{y})}{\overline{x^{2}}-(\overline{x})^{2}}$$

```{r calculate_sums_and_means, echo=FALSE}

augmented_df <- data.frame(x=x,y=y,xy=x*y,x_sq=x^2) 
knitr::kable(augmented_df, col.names=c("x","y","xy","$x^{2}$"))

sum_row <- as.data.frame(lapply(augmented_df,sum))
knitr::kable(sum_row, col.names = c("$\\sum{x}$","$\\sum{y}$","$\\sum{xy}$","$\\sum{x^{2}}$"))

mean_row <- as.data.frame(lapply(augmented_df,mean))
knitr::kable(mean_row, col.names = c("$\\overline{x}$","$\\overline{y}$","$\\overline{xy}$","$\\overline{x^{2}}$"))

m_numerator <- mean_row$xy - (mean_row$x * mean_row$y)
m_denominator <- mean_row$x_sq - mean_row$x^2
m <- m_numerator/m_denominator

b <- mean_row$y - (m * mean_row$x)
```

m = (`r mean_row$xy` - (`r mean_row$x`)(`r mean_row$y`))/(`r mean_row$x_sq` - (`r mean_row$x`)^2)  
m = (`r mean_row$xy` -  &nbsp;&nbsp;  `r mean_row$x * mean_row$y` &nbsp; )/(`r mean_row$x_sq` - `r mean_row$x ^ 2`)  
m = `r m_numerator`/`r m_denominator`  

m = `r m`

And for

$$b=\overline{y} - m\overline{x}$$

b = `r mean_row$y` - `r m` (`r mean_row$x`)  
b = `r b`

Thus giving the line y = `r m`x + `r b`, which we'll plot over the points.

\newpage

Remember that our "manually calculated" regression formula is 

y = `r m`x + `r b`

(m = `r m`, b = `r b`)  


```{r plot_with_line}

# plot the points
plot(x,y)

# capture the lowest and highest x values so that our regression line will span all values have the same length
ideal_x <- c(min(x),max(x))

# calculate 
ideal_y <- lapply(ideal_x, function(x) {m*x + b}) # lapply "applies" our formula to get the model's y coordinates

# Draw the regression line
lines(ideal_x,ideal_y)
```

Or if you want to take all the fun out of it and just use R functions to do the work:

```{r using_r_functions}

# Linear regression model for our points
my_model <- lm(y ~ x)

# Plot the points
plot(x,y)

# Plot the regression line for our model
lines(x,predict(my_model))
```

